Data 621---HW1---Regression Model---Version 2

### 2020-09-09 HW1/RegModel_Adler.Rmd
Finished part 3 on part 4

# Data Exploration
## Task
Describe the size and the variables in the moneyballtraining data set. Consider
that too much detail will cause a manager to lose interest while too little
detail will make the manager consider that you aren’t doing your job. Some
suggestions are given below. Please do NOT treat this as a check list of things
to do to complete the assignment. You should have your own thoughts on what to
tell the boss. These are just ideas.

@@

Our suggestion would be to use only `TEAM_PITCHING_HR` as that has a slightly
higher correlation with target wins, and leave all the other variables in the
analysis.

@@

# Remove TEAM_BATTING_HR

MBmelt <- MBmelt[Metric != 'TEAM_BATTING_HR']

@@

Unfortunately, the scaling introduces infinities, as most of the Box-Cox lamdas
are non-positive, which transforms 0 into infinity. These were converted to NAs
for the purposes of imputation, after which scaling and centering will be done
again. This is the danger of too many transforms. Each one induces a "stress" on
the relationship between the independent and dependent variables. At what point
does that relationship fracture?

@@

## Analysis
Normally, we would have thrown everything into `caret`, but for the purposes of
this homework, we will use calls to base R for the null and saturated models and
the `stepAIC` function from the `MASS` package to perform a stepwise linear
regression using AIC as our goodness of fit metric.

@@

We will show three models: the null model, the saturated model, and the model
resulting from the stepwise regression. For the regression, we will consider at
most pairwise interactions within the classes of metrics.

@@

nullModel <- lm(TARGET_WINS ~ 1, data = MBsc)
snm <- summary(nullModel)

saturatedModel <- lm(TARGET_WINS ~ ., data = MBsc)
ssm <- summary(saturatedModel)

stepModel <- stepAIC(saturatedModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
   TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_SO) ^ 2 + (TEAM_FIELDING_DP +
   TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB + TEAM_PITCHING_H +
   TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1), direction = 'both')

sstm <- summary(stepModel)
```
```{r nullOutput}
knitr::kable(snm$coefficients, digits = 3L,
             caption = 'Null Model Output')
```
The null model has an adjusted \(R^2\) of
`r prettyNum(snm$adj.r.squared, digits = 3L)` and an AIC of
`r prettyNum(AIC(nullModel), digits = 2L, big.mark = ',')`.

```{r satOutput}
knitr::kable(ssm$coefficients, digits = 3L,
             caption = 'Saturated Model Output')
```
The saturated model has an adjusted \(R^2\) of
`r prettyNum(ssm$adj.r.squared, digits = 3L)` and an AIC of
`r prettyNum(AIC(saturatedModel), digits = 2L, big.mark = ',')`.

@@

The step model allowing pairwise interactions between the four classes of
metric: base running, batting, fielding, and pitching, has an adjusted \(R^2\)
of `r prettyNum(sstm$adj.r.squared, digits = 3L)` and an AIC of
`r prettyNum(AIC(stepModel), digits = 2L, big.mark = ',')`.

@@

 * Friendly, Michael (2002) Corrgrams: Exploratory displays for correlation
 matrices, *The American Statistician*, 56:4, 316-324, DOI: 10.1198/000313002533,
 http://euclid.psych.yorku.ca/datavis/papers/corrgram.pdf

>>>>

# Data Exploration
## Task
Describe the size and the variables in the moneyball training data set. Consider
that too much detail will cause a manager to lose interest while too little
detail will make the manager consider that you aren’t doing your job. Some
suggestions are given below. Please do NOT treat this as a check list of things
to do to complete the assignment. You should have your own thoughts on what to
tell the boss. These are just ideas.

@@

Initially, our thought was to only use `TEAM_PITCHING_HR` as that has a slightly
higher correlation with target wins, but after running the model, it was clear
that there was valid information captured in both variables so both were left in.

@@

@@

Unfortunately, the scaling introduces infinities, as most of the Box-Cox lambdas
are non-positive, which transforms 0 into infinity. These were converted to NAs
for the purposes of imputation, after which scaling and centering will be done
again. This is the danger of too many transforms. Each one induces a "stress" on
the relationship between the independent and dependent variables. At what point
does that relationship fracture?

@@

## Analysis
Normally, we would use a statistical packahe like `caret`, which not only
simplifies model construction but also takes care of cross-validation. For the
purposes of this homework, and to demonstrate understanding of the underlying
decisions involved in predictive modeling, we will use calls to base R for the
null and saturated models and the `stepAIC` function from the `MASS` package to
perform a stepwise linear regression using AIC as our goodness of fit metric.
In the final section, we may pivot to using `caret` for final model selection.

@@

### Linear Models
We will show three models: the null model, the saturated model, and the model
resulting from the stepwise regression. For the regression, we will consider at
most pairwise interactions within the classes of metrics.

@@

nullModel <- lm(TARGET_WINS ~ 1, data = MBsc)
snm <- summary(nullModel)
nmr2 <- prettyNum(snm$r.squared, digits = 3L)
nmar2 <- prettyNum(snm$adj.r.squared, digits = 3L)
nmAIC <- prettyNum(AIC(nullModel), digits = 2L, big.mark = ',')

saturatedModel <- lm(TARGET_WINS ~ ., data = MBsc)
ssm <- summary(saturatedModel)
smr2 <- prettyNum(ssm$r.squared, digits = 3L)
smar2 <- prettyNum(ssm$adj.r.squared, digits = 3L)
smAIC <- prettyNum(AIC(saturatedModel), digits = 2L, big.mark = ',')

stepModel <- stepAIC(saturatedModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1),
  direction = 'both')

sstm <- summary(stepModel)
stmr2 <- prettyNum(sstm$r.squared, digits = 3L)
stmar2 <- prettyNum(sstm$adj.r.squared, digits = 3L)
stmAIC <- prettyNum(AIC(stepModel), digits = 2L, big.mark = ',')

```
```{r nullOutput}
knitr::kable(snm$coefficients, digits = 3L,
             caption = 'Null Model Output')
```
The null model has an \(R^2\) of `r nmr2`, an adjusted \(R^2\) of `r nmar2` and
an AIC of `r nmAIC`.

```{r satOutput}
knitr::kable(ssm$coefficients, digits = 3L,
             caption = 'Saturated Model Output')
```
The saturated model has an \(R^2\) of `r smr2`, an adjusted \(R^2\) of `r smar2`
and an AIC of `r smAIC`.

@@

The step model allowing pairwise interactions within the four classes of
metric: base running, batting, fielding, and pitching, has an \(R^2\) of
`r stmr2`, an adjusted \(R^2\) of `r stmar2` and an AIC of `r stmAIC`.

### Discussion
The first item for discussion is the observation that the intercept does **not**
change between the null (intercept-only) model and the saturated (all variables
included) model. There are 162 games in a baseball season. According to both the
null and saturated models, given no other information, a team will *lose*
slightly more than half its games. Whereas in the stepwise model allowing for
pairwise interactions within classes, given no other information, a team will
*win* slightly more than half of its games. Apparently, inteactions unearth
somewhat more information about why teams **lose** rather than why they **win**.

The stepwise model has the best \(R^2\) of the three and the best AIC. Using
accepted rules of thumb, a difference of more than 10 indicates no support for
the model with the higher AIC. The magnitude is irrelevant; it is a function of
the number of observations. The difference between the AIC of the saturated
model and the step model is
`r prettyNum(AIC(saturatedModel) - AIC(stepModel), digits = 2L)`. This shows
that the saturated model, and obviously the null model, should be completely
ignored in favor of the stepwise model with pairwise within class interactions.

Most of the coefficients are reasonable within the context of the game of
baseball. These two statements are axiomatic:

  * The only way to win is to have the higher score at the end of the game.
  * The only way to score is for a baserunner to cross home plate.

With those in mind, we can make the following observations about the linear
non-interactive coefficients.

 * Reasonable
   * Caught stealing removes baserunners; negative coefficient makes sense.
   * Stolen bases get a runner closer to home plate; positive coefficient makes
   sense.
   * Triples get a runner very close to home plate; positive coefficient makes
   sense.
   * Walks are free baserunners; positive coefficient makes sense.
   * Hits increase the number of baserunners; positive coefficient makes sense.
   * Hitting home runs directly increase the score; positive coefficient makes
   sense.
   * Striking out reduces the number of baserunners; negative coefficient makes
   sense.
   * Errors allow the other team free baserunners; negative coefficient makes
   sense.
   * Giving up home runs gives the opponent scores; negative coefficient makes
   sense.
   * Getting strikeouts reduces the opponents baserunners; positive coefficient
   makes sense.
 * Curious
   * Hitting doubles get runners on base; why negative?
   * Turning double plays reduce baserunners; why negative?
   * Giving up walks allows the other team free baserunners; its coefficient is
   both insignificant and near 0.
   * Giving up hits allows the other team baserunners; its coefficient is
   both insignificant and near 0.

An absolutely fascinating observation. In the first draft of this exercise,
`TEAM_BATTING_HR` was removed due to its high correlation with
`TEAM_PITCHING_HR`. Note the coefficients, batting is +10 and pitching is -7.
However, this is such as strong indicator, that in the first run, pitching was
given a *postive* coefficient of around +3. In hindsight this is because
*it was being used as an indicator for* ***batting!!*** The correlation allowed
the use of pitching HRs as an indicator for the hidden batting HRs! Once both
were restored to the model, the logical coefficients surfaced. Another reason
why models should not be trusted out of the box, but all model results should
be reviewed for sanity and sense!!

The curiosities above can *possibly* be resolved by looking at the interaction
terms. When giving up walks is interacted with strikeouts,
`TEAM_PITCHING_BB:TEAM_PITCHING_SO`, the coefficent is positive and significant.
Perhaps pitchers with many strikeouts also walk a lot of batters so giving up
walks on its own is less relevant. Similarly, the interaction between giving
up hits and strikeouts, `TEAM_PITCHING_H:TEAM_PITCHING_SO `, is highly negative
and significant. Once again, it probably captures most of the giving up hits
information making the singleton less relevant.

A possible explanation for the negative coefficent for getting double plays, is
that double plays require at least two people on base. That means that the
oppenent has a lot of base runners, which is very highly correlated with
scoring.

The behavior of doubles remains confusing. Unless its hiding something like
a team's propensity to strand runners on base. It would be interesting to see
a breakdown between the American and National leagues, as the latter tends to be
somewhat better at "small ball" and moving the runners along.

# Build Models
## Task
Decide on the criteria for selecting the best multiple linear regression model.
Will you select a model with slightly worse performance if it makes more sense
or is more parsimonious? Discuss why you selected your model. For the multiple
linear regression model, will you use a metric such as Adjusted \(R^2\), RMSE,
etc.? Be sure to explain how you can make inferences from the model,
discuss multi-collinearity issues (if any), and discuss other relevant model
output.

Using the training data set, evaluate the multiple linear regression model based
on (a) mean squared error, (b) \(R^2\), (c) F-statistic, and (d) residual plots.
Make predictions using the evaluation data set.

## Analysis
At this point, to properly fulfill this portion, cross-validation should be used
in model selection. That is not possible here given the model selection above.
That being said, given that the purpose of this exercise is **not** to
demonstrate facility with tuning, the three models developed above will be used
as the representatives.

### Criteria
As a devout follower of Burnham & Anderson (2002), I prefer to use AIC whenever
possible. This includes inherent recognition of parsimony. As discussed in the
previous section, for an AIC difference of 10 or greater, the model with the
higher AIC has effectively no support. The stepwise model with within-class
pairwise interactions had the lowest AIC by hundreds.

@@

 * Burnham, K.P. and Anderson, D.R. (2002) "Model Selection and Inference: A
 Practical Information-Theoretic Approach". 2nd Edition, Springer-Verlag,
 New York. http://dx.doi.org/10.1007/b97636
 * Friendly, Michael (2002) Corrgrams: Exploratory displays for correlation
 matrices, *The American Statistician*, 56:4, 316-324, DOI: 10.1198/000313002533,
 http://euclid.psych.yorku.ca/datavis/papers/corrgram.pdf
